Composable data visualizations

### 2020-05-11 wip-compost/paper.tex
Interactivity

\section{Interactive charts}

@@

>>>>

\section{Interactive charts: Domain-specific event handling}
Data visualizations published on the web often also feature interactivity. Standard forms of
interactivity include animations, hover labels or zooming. More interesting custom forms
include the ``You draw It'' visualization introduced by the New York Times. % cite
The chart shows only the first half of the data, such as a timeline, and the reader has to
guess the second half before clicking a button and seeing the actual data. Standard forms of
interactivity are supported by high-level libraries. Google Charts supports panning using
drag \& drop, zooming to a selected chart range and animations. Custom interactivity can be
implemented using low-level libraries such as D3, but doing so requires directly handling
JavaScript events.

\end{figure}
\begin{figure}
\begin{equation*}
\begin{array}{rclcl}
s & = & \kvd{mouseUp}~(\lambda \,x~y \rightarrow e), s  &|& \kvd{mouseDown}~(\lambda \,x~y\rightarrow e), s  \\
    &|& \kvd{mouseMove}~(\lambda \,x~y \rightarrow e), s &|& (\ldots)
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{Additional combinators for mouse-based interaction, extending earlier definition of $s$.}
\label{fig:mouse}
\end{figure}

Compost can be used in conjunction with a virtual-dom library to create interactive data
visualizations based on the Elm architecture. An interactive visualization is described using
a pair of functions -- the \emph{view} function creates a chart based on the current state and
the \emph{update} function specifies how the state changes when an event occurs.

To support handling of mouse-based events, Compost adds three additional primitives to the definition
of shape, shown in Figure~\ref{fig:mouse}. The three new primitives make it possible to handle
three common mouse events using custom functions $\lambda \;x~y\rightarrow e$, specified in the
host language. The most interesting aspect is that the functions are given X and Y coordinates
of the event specified in the domain of the chart. This means that if the user clicks on the
bar representing the Conservative party in a bar chart, the values might be, for example,
$\kvd{cat}~\strf{Conservative},\num{0.75}$ for X and $\kvd{cont}~\num{120.5}$ for Y.

To illustrate building of interactive data visualizations using Compost, we look at one aspect of
the ``You draw It'' visualizations. We want to create a bar chart where the user can use drag \& drop
to move individual bars. The resulting chart can be seen in Figure~\ref{fig:youdraw}.
The first step is to define types representing the state and events that can occur:
%
\begin{equation*}
\begin{array}{lll}
\fkvd{type}~\ident{State} &\narrow{=}& \ident{bool}~*~(\ident{string}~*~\ident{int})~\ident{list}\\[0.5em]
\fkvd{type}~\ident{Event} &\narrow{=}& \ident{Update}~\fkvd{of}~(\ident{string}~*~\ident{int})\\
                          &\;\narrow{|}& \ident{Moving}~\fkvd{of}~\ident{bool}
\end{array}
\end{equation*}

yadda

\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{update}~(\fkvd{true},s)~(\ident{Update}(p, v)) =
  \fkvd{true}, \ident{map}~(\lambda(k, o) \rightarrow k, \fkvd{if}~k = p~\fkvd{then}~v~\fkvd{else}~o)~s\\
\hspace{1.05em}~\ident{update}~(m,s)~(\ident{Update}(\_, \_)) = m,s\\
\hspace{1.05em}~\ident{update}~(\_,s)~(\ident{Moving}(m)) = m,s\\
\end{array}
\end{equation*}

yadda

\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{render}~t~(m, s)~=\\
\quad \kvd{axis}_l~(\kvd{axis}_b~(\kvd{explicitScale}_y~(\kvd{continuous}~\num{0},\num{400}),\\
\qquad (~\kvd{mouseMove}~(\lambda \,(\kvd{cat}~p,\_)~(\kvd{cont}~v)\rightarrow \ident{trigger}(\ident{Update}(p, v))),\\
\qquad (~\kvd{mouseUp}~(\lambda \,\_~\_\rightarrow \ident{trigger}(\ident{Moving}(\fkvd{true}))),\\
\qquad (~\kvd{mouseDown}~(\lambda \,\_~\_\rightarrow \ident{trigger}(\ident{Moving}(\fkvd{false}))),~\kvd{overlay}~[\\
\qquad \qquad \fkvd{for}~\ident{party}, \ident{mps}~\fkvd{in}~s~\rightarrow~\kvd{padding}~0,10,0,10,~\kvd{fill}~(\ident{color}~\ident{party}),\\
\qquad \qquad\quad\quad [\,(\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\ident{mp}),\\
\qquad \qquad\quad\quad \hspace{0.38em}(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mps}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}) \,])~~])))))\\
\end{array}
\end{equation*}

@@

\begin{figure}[h]
  \includegraphics[scale=0.57]{figures/youdraw}
  \vspace{0.25em}
  \caption{You draw.}
  \label{fig:youdraw}
\end{figure}
